Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-02-03 paper/paper.tex
fleshed out mnist section

% general intro text with motivation
Deep neural networks are an integral part of machine learning and data science toolset
for practical data-driven problem solving. With continual miniaturization ever more
applications can be found in embedded systems. Common embedded applications include
on-device image recognition and signal processing. Despite recent advances in generalization
and optimization theory specific to deep networks, deploying in actual embedded hardware
remains a challenge due to storage, real-time throughput, and arithmetic (computational)
complexity restrictions \citep{he_amc:_2018}. Therefore compression methods for achieving
high model sparsity and numerical efficiency without losing much in performance are
especially relevant.
% Solving the storage and artihmetic constraints encourages development of model compression
% and sparsification methods, that focus on favourable tradeoff between performance and size.

@@

In this section we study the complex variational dropout techniques discussed above
on the tasks and datasets, that were studied in prior research on complex-valued networks.
All experiments share the following settings unless specified otherwise:
\begin{itemize}
  \item ADAM optimizer with the base learning rate of $10^{-3}$
  \item learning rate the annealing schedule from \citep{trabelsi_deep_2017}
  \item global gradient clipping within an $\ell_2$-norm ball of radius $0.05$
\end{itemize}
Every experiment is replicated several times to take random effects into account, e.g.
random network initialization prior to training, randomly ordered batches during SGD.
Throughout every experiment on MNIST-like datasets (sec.~\ref{sub:mnist_like_datasets},
we \textit{fixed} a random subset of the train split of each dataset, and used the test
split to gauge the final performance. The same approach was applied to CIFAR10 (sec.~%
\ref{sub:cifar10_100} experiments, which could partly explain the overall poor test
performance we got from the tested VGG16 model ($\real$ and $\cplx$ version).

@@

The goal of the following experiments to study the sparsity-performance trade-off of the
proposed Bayesian sparsification methods for $\cplx$-valued networks. To this end we employ
the following training plan in every experiment. The model is trained in three successive
stages: ``dense'' $\to$ ``sparsify'' $\to$ ``fine-tune''. Parameter values and the internal
state of the optimizer for related parameters are transferred between stages.

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-bayesian architecture and using only the likelihood term from \eqref{eq:elbo} without
variational dropout.
%
During the \textit{``sparsify''} stage we make every layer bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:variational_dropout} (or sec.~\ref{ssub:empirical_bayes}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,. \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.
%
The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks. The test
performance is reported for the model at the start and termination of the final stage.

Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the variational approximation factorizes into univariate distributions, whether a $\cplx$
or $\real$ parameter is zero or not is determined from individual relevance scores $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$: non-zero if $\log \alpha \leq \tau$, zero otherwise. Figure~\ref{fig:log-relevance}
(top) displays the general shape of the distribution of $\log \alpha$ values from all
experiments in sec.~\ref{sub:mnist_like_datasets}. In fact, this appears to be the typical
in every experiment conducted in this study: there always is a single mode within the
moderately negative range, and one or two modes to the right of zero. The relative mass
of each mode changes with $C$ and the higher the coefficient is, the more mass is shifted
to the right. In every experiment the trough between the peaks of the negative mode and
the first positive one is around zero.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth]{../assets/log-alpha__sum__mnist__20200131-202713.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth]{../assets/log-alpha__sum__fashion-mnist__20200131-205926.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    \textit{top}: Estimate of $\log\alpha$ density for a convolutional and dense layer
    from the convolutional model of sec.~\ref{sub:mnist_like_datasets}.
    \textit{bottom}: The probability of a parameter being within $\pm \delta$ relative
    tolerance of its mean.
  }
  \label{fig:log-relevance}
\end{figure}

We decide if a parameter is \textit{``non-zero''} if it is within $\delta$ relative
tolerance of its mean with high probability under the approximate posterior. Since
\eqref{eq:c-gauss-vi} is Gaussian, $
  \tfrac{\lvert w - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with shape $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    = \Pr\biggl(
      \frac{\lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
        \leq \frac{\delta^2}{\alpha}
    \biggr)
    = \chi^2_k\biggl(\frac{\delta^2}{\alpha}\biggr)
  \,.
\end{equation}
Figure~\ref{fig:log-relevance} (bottom) shows the probability of a $\real$ (or $\cplx$) 
parameter being non-zero for various $\delta\in(0, \tfrac12)$ as a function of $\log \alpha$.
For a reasonably generous $\delta = 25\%$ tolerance all values of $\log \alpha$ below
$-2.5$ yield a relatively high chance of a non-zero parameter. Given that $q_\theta$
is merely an approximation, we opt for a somewhat higher $\log \alpha$ threshold, and
pick $\tau = -\tfrac12$, which is much lower than $\tau = 3$, recommended by
\citet{molchanov_variational_2017}. Additional rationale behind this decision is the
extra fine tuning, which trains the compressed model.

@@

We conduct a moderately sized numerical experiment on MNIST and similar $28\times 28$
greyscale image datasets to see whether the proposed $\cplx$-variational dropout indeed
sparsifies complex-valued networks. We compare with the studied $\cplx$-valued networks
against their $\real$-valued counterparts both in terms of the resulting sparsity and
test performance. For a fair comparison, we make the $\cplx$-valued network have
approximately half as many parameters, \citep{monning_evaluation_2018}.

@@

% a couple of plots from MNIST and Fashion-MNIST
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.95\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth]{../assets/sum__mnist__fine-tune.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{0.95\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth]{../assets/sum__fashion-mnist__fine-tune.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    The accuracy-compression tradeoff after ``fine-tune''.
    \textit{top}: MNIST \citep{citation_needed};
    \textit{bottom}: Fashion MNIST \citep{citation_needed}.
  }
  \label{fig:mnist-like_tradeoff}
\end{figure}

% write a note on the incompressible parameters (bias, batch norm etc.)
% discuss the plots describing how the test-performance grows with compression and then drops if the model is too sparse
% hinting at overfitting/regularization effects of masking and sparsification

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-fine-tune.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v2-fine-tune.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    Average Percision frontier for the convolutional model of \citet{trabelsi_deep_2017}.
    Shaded horiontal region represents the $\mu\pm\sigma$ bands for the performance of the
    uncompressed model.
    \textit{top}: Original model ($k=6$) compressed with \textbf{\color{orange} VD}
    (\ref{ssub:variational_dropout}) and \textbf{\color{blue} ARD} (\ref{ssub:empirical_bayes});
    \textit{bottom}: The model of \citep{trabelsi_deep_2017} with kernel size $3$
    compressed with \textbf{\color{blue} VD} (\ref{ssub:variational_dropout}).
  }
  \label{fig:musicnet_tradeoff}
\end{figure}

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-c-ard-tradeoff.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-c-ard-tradeoff.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    Pre ``finte-tune'' performance of the best models from figure \ref{fig:musicnet_tradeoff}
    compressed with methods of sec.~\ref{sec:dropout_for_c_valued_layers}:
    (\ref{ssub:variational_dropout} -- \textit{top}, \ref{ssub:empirical_bayes} --  \textit{bottom}).
  }
  \label{fig:musicnet_tradeoff_pre_fune-tune}
\end{figure}
% subsection muscinet (end)

>>>>

% general intro text with motivation
Deep neural networks are an integral part of machine learning and data science toolset
for practical data-driven problem solving. With continual miniaturization ever more
applications can be found in embedded systems. Common embedded applications include
on-device image recognition and signal processing. Despite recent advances in generalization
and optimization theory specific to deep networks, deploying in actual embedded hardware
remains a challenge due to storage, real-time throughput, and arithmetic (computational)
complexity restrictions \citep{he_amc:_2018}. Therefore compression methods for achieving
high model sparsity and numerical efficiency without losing much in performance are
especially relevant.
% Solving the storage and artihmetic constraints encourages development of model compression
% and sparsification methods, that focus on favourable trade-off between performance and size.

@@

In this section we study the complex variational dropout techniques discussed above
on the tasks and datasets, that were studied in prior research on complex-valued networks.

@@

The goal of this section to study the compression-performance trade-off of the proposed Bayesian
sparsification methods for $\cplx$-valued networks. To this end we employ the following training
plan in every experiment. The model is trained in three successive stages: ``dense'' $\to$
``sparsify'' $\to$ ``fine-tune''. Parameter values and the internal state of the optimizers
are transferred between stages.

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is computed based on the number of floating point values needed
to specify the model and equals $
  \frac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros after sparsification, $n_\mathtt{par}$
is the total number of floats. Thus each parameter is counted once in a $\real$-valued
network, but twice in a $\cplx$-network, being a pair of real-imaginary floats. Each model
has incompressible parameters, such as layer biases, and shifts and scalings from $\real$-
and $\cplx$-batch normalization layers, if present.

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-Bayesian architecture and using only the likelihood term from \eqref{eq:elbo} without
variational dropout.
%
During the \textit{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:variational_dropout} (or sec.~\ref{ssub:empirical_bayes}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,. \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.
%
The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks. The test
performance is reported for the model at the start and termination of the final stage.

Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the variational approximation factorizes into univariate distributions, whether a $\cplx$
or $\real$ parameter is zero or not is determined from individual relevance scores $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$: non-zero if $\log \alpha \leq \tau$, zero otherwise. Figure~\ref{fig:log-alpha__mnist-like}
displays the general shape of the distribution of $\log \alpha$ on the Fashion-MNIST dataset
in sec.~\ref{sub:mnist_like_datasets}. This shape appears to be typical in every experiment
conducted in this study: there always is a single mode within the moderately negative
range, and one or two modes to the right of zero. The relative mass of each mode changes
with $C$ and the higher the coefficient is, the more mass is shifted to the right. In every
experiment the trough between the peaks of the negative mode and the first positive one
is around zero.

\begin{figure}[ht]
  \centering
  \includegraphics[draft,width=1.\linewidth]{../assets/log-alpha__sum__fashion-mnist.pdf}
  \caption{%
    Smoothed density of $\log\alpha$ for a convolutional and dense layer from the
    model in sec.~\ref{sub:mnist_like_datasets}.
  }
  \label{fig:log-alpha__mnist-like}
\end{figure}

We decide if a parameter is \textit{``non-zero''} if it is within $\delta$ relative
tolerance of its mean with high probability under the approximate posterior. Since
\eqref{eq:c-gauss-vi} is Gaussian, $
  \tfrac{\lvert w - \mu \rvert^2}
        {\alpha \lvert \mu \rvert^2}
$ is $\chi^2_k$ distributed with shape $k=1$ ($\real$) or $2$ ($\cplx$), whence
\begin{equation} \label{eq:prob_relevant}
  \Pr\bigl(
    \lvert w - \mu \rvert \leq \delta \lvert \mu \rvert
  \bigr)
    = \Pr\biggl(
      \frac{\lvert w - \mu \rvert^2}{\alpha \lvert \mu \rvert^2}
        \leq \frac{\delta^2}{\alpha}
    \biggr)
    = \chi^2_k\biggl(\frac{\delta^2}{\alpha}\biggr)
  \,.
\end{equation}
% Figure~\ref{fig:log-alpha__mnist-like} (bottom) shows the probability of a $\real$
% (or $\cplx$)  parameter being non-zero for various $\delta\in(0, \tfrac12)$ as a function
% of $\log \alpha$.
For a reasonably generous $\delta = 25\%$ tolerance all values of $\log \alpha$ below
$-2.5$ yield a relatively high chance of a non-zero parameter. Given that $q_\theta$
is merely an approximation, we opt for a somewhat higher $\log \alpha$ threshold, and
pick $\tau = -\tfrac12$, which is much lower than $\tau = 3$, recommended by
\citet{molchanov_variational_2017}. Additional rationale behind this decision is the
extra fine tuning, which trains the compressed model.

% common experiment settings
Every experiment is replicated several times to take random effects into account, e.g.
random network initialization prior to training, randomly ordered batches during SGD.
Unless specified otherwise, every experiment uses global gradient clipping within an
$\ell_2$-norm ball of radius $0.5$, runs SGD with ADAM optimizer with the base learning
rate of $10^{-3}$ which is scaled after $5$, $10$, and $20$ epochs by $\tfrac1{10}$,
$\tfrac1{20}$ and $\tfrac1{100}$, respectively.

@@

We conduct a moderately sized numerical experiment on MNIST and similar $28\times 28$
greyscale image datasets to study the compression-performance trade-off of the proposed
$\cplx$-variational dropout on simple architectures. We use classification accuracy as
the target performance metric. We compare with the $\real$-valued networks against their
$\cplx$-valued counterparts with approximately half as many parameters, as suggested
by \citet{monning_evaluation_2018}.

% dataset size
We deliberately use a {fixed} random subset of $10k$ images from the train split of each
dataset to fit the networks, and gauge the final performance on the usual test split.
Limiting the size was done primarily to test the strength of regularization properties
of variational dropout, and to make experiments quicker.
% image features
Since image data is not naturally $\cplx$-valued, we reprocess images by either embedding
$\real$-data into $\cplx$ assuming $\Re z = 0$ (\texttt{raw}), or applying the two-dimensional
Fourier Transform (\texttt{fft}), shifting lower frequencies to the centre of the image. We
do not investigate the option, proposed by \citet{trabelsi_deep_2017}, to train an auxiliary
model to synthesize the imaginary component from the real data.

% model architectures
We study the following $\real$/$\cplx$ networks:
\begin{itemize}
  \item \textit{SimpleConvModel}: $2d$ convnet with Average pooling ($2\times 2$)
  with layers $c(n_1, 5, 1) \to c(n_2, 5, 1) \to d(n_3) \to d(n_\mathtt{out})$
  \item \textit{TwoLayerDenseModel}: wide shallow network with structure
  $d(n_1) \to d(n_\mathtt{out})$
\end{itemize}
where $c(n, k, s)$ denotes a convolution with $n$ output channels, kernel width $k$ and
stride $s$, and $d(n)$ is a dense layer with $n$ outputs. The values of $n_\cdot$ depend
on the input features and are summarized in table~\ref{tbl:mnist-like-model-spec}.
\begin{table}[hb]
  \centering
  \caption{Architecture parameters for models studied on MNIST-like datasets.}
  \label{tbl:mnist-like-model-spec}
  \begin{tabular}{lrr}
    \toprule
    $(n_1, n_2, n_3)$  & $\real$-\textit{SimpleConvModel} & $\cplx$-\textit{SimpleConvModel} \\
    \midrule
    (\texttt{raw}) & $20, 50, 500$ & $10, 25, 250$ \\
    (\texttt{fft}) & $40, 100, 1000$ & $20, 50, 500$ \\
    \bottomrule
  \end{tabular}
  \begin{tabular}{lrr}
    \toprule
    $(n_1)$  & $\real$-\textit{TwoLayerDenseModel} & $\cplx$-\textit{TwoLayerDenseModel} \\
    \midrule
    (\texttt{raw}) & $4096$ & $2048$ \\
    (\texttt{fft}) & $8192$ & $4096$ \\
    \bottomrule
  \end{tabular}
\end{table}

% discuss the plots describing how the test-performance grows with compression and then drops if the model is too sparse

% hinting at overfitting/regularization effects of masking and sparsification

% a couple of plots from MNIST and Fashion-MNIST
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.95\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/sum__mnist__fine-tune.pdf}
    % \caption{} \label{fig:log-alpha-kde}
  \end{subfigure} \\%
  % \hfill
  \begin{subfigure}[b]{0.95\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/sum__fashion-mnist__fine-tune.pdf}
    % \caption{} \label{fig:log-alpha-proba}
  \end{subfigure}
  \caption{%
    The accuracy-compression trade-off after ``fine-tune''.
    \textit{top}: MNIST \citep{citation_needed};
    \textit{bottom}: Fashion MNIST \citep{citation_needed}.
  }
  \label{fig:mnist-like_trade-off}
\end{figure}

The same approach was applied to
CIFAR10 (sec.~\ref{sub:cifar10_100} experiments, which could partly explain the overall
poor test performance we got from the tested VGG16 model ($\real$ and $\cplx$ version).

% comparison strategy
We conduct a moderately sized numerical experiment on MNIST and similar $28\times 28$
greyscale image datasets to see whether the proposed $\cplx$-variational dropout indeed
sparsifies complex-valued networks. We compare with the studied $\cplx$-valued networks
against their $\real$-valued counterparts both in terms of the resulting sparsity and
test performance. For a fair comparison, we make the $\cplx$-valued network have
approximately half as many parameters, \citep{monning_evaluation_2018}.

@@

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-fine-tune.pdf}
    % \caption{} \label{fig:log-alpha-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v2-fine-tune.pdf}
    % \caption{} \label{fig:log-alpha-proba}
  \end{subfigure}
  \caption{%
    Average Percision frontier for the convolutional model of \citet{trabelsi_deep_2017}.
    Shaded horiontal region represents the $\mu\pm\sigma$ bands for the performance of the
    uncompressed model.
    \textit{top}: Original model ($k=6$) compressed with \textbf{\color{orange} VD}
    (\ref{ssub:variational_dropout}) and \textbf{\color{blue} ARD} (\ref{ssub:empirical_bayes});
    \textit{bottom}: The model of \citep{trabelsi_deep_2017} with kernel size $3$
    compressed with \textbf{\color{blue} VD} (\ref{ssub:variational_dropout}).
  }
  \label{fig:musicnet_trade-off}
\end{figure}

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-c-ard-trade-off.pdf}
    % \caption{} \label{fig:log-alpha-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[draft,width=1.\linewidth]{../assets/grid_cplx_fine_kl_div_v3_fast-c-ard-trade-off.pdf}
    % \caption{} \label{fig:log-alpha-proba}
  \end{subfigure}
  \caption{%
    Pre ``finte-tune'' performance of the best models from figure \ref{fig:musicnet_trade-off}
    compressed with methods of sec.~\ref{sec:dropout_for_c_valued_layers}:
    (\ref{ssub:variational_dropout} -- \textit{top}, \ref{ssub:empirical_bayes} --  \textit{bottom}).
  }
  \label{fig:musicnet_trade-off_pre_fune-tune}
\end{figure}
% subsection muscinet (end)
